---
title: "An Introduction to the multiGSEA Package"
author: "Steve Lianoglou"
date: "`r Sys.Date()`"
abstract: >
  The *multiGSEA* package facilitates the running and comparison of many GSEA
  approaches for a single experimental contrast. This vignette covers the
  features found in *multiGSEA (>= 0.5.5)*, which is currently available in
  `apps/R/current`, and soon to `be apps/R/stable` (ie. Bioc 3.3).
vignette: >
  %\VignetteIndexEntry{An Introduction to the multiGSEA Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
# date: "r doc_date()"
# package: "r pkg_ver('multiGSEA')"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
      smooth_scroll: yes
    css: mystyles.css
---

```{r init, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, warning=FALSE, message=FALSE, error=FALSE, dpi=150)
```

# Overview

The `multiGSEA` package was built to facilitate the comparison of multiple
GSEA methods over given experimental contrast. The outputs of the various
methods are interanlly saved in their original form and can be retrieved
for individual scrutiny, or they can be combined to facilitate comparison.

The package also provides a shiny application that allows the user to
interactively explore the gene set membership and enrichment.

## Background

The initial GSEA methods that multiGSEA wrapped were the ones provided by limma
and edgeR. As such, most analyses using multiGSEA expect you to re-use the same
data objects used for differential expression analysis, namely:

* Expression data (an `EList`, `DGEList`, or expression matrix)
* A design matrix
* A contrast vector/matrix (if your design and comparison require it)

Currenty supported gene set enrichment methods include:

  * camera
  * (m)roast / fry
  * romer
  * geneSetTest
  * goseq
  * hypergeometric test

# Standard Workflow

We'll use the `"r Biocpkg("airway")"` data package from [Himes et al.][himes] as
the basis for our analysis here[^rnaseqworkflow]. This dataset contains
treatment (with the glucocorticoid dexamethasone) and control samples from four
airway smooth muscle (ASM) cell lines in order to study the mechanism by which
glucocorticoids suppress inflammation in ASM.

[himes]: http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0099625

[^rnaseqworkflow]: This dataset is also used as the basis for the RNA-seq
workflow tutorial published by Love et al. on
[f1000research](http://f1000research.com/articles/4-1070/v1)


## Data Setup

multiGSEA is most straightforward to use when our data objects and analysis are
performed with either the edgeR or voom/limma pipelines. In order to facilitate
our analyses, we first convert the `airway` `SummarizedExperiment` to a
`DGEList` and add simple gene annotations to the data.

```{r data-setup, eval=!exists('y.all')}
library(org.Hs.eg.db) ## To annotate ENSGIDs
library(edgeR)
library(airway)
library(magrittr)
library(dplyr)
library(dtplyr)
library(ggplot2)
theme_set(theme_bw())

data("airway")
se <- airway
se$dex <- relevel(se$dex, 'untrt')
y.all <- DGEList(
  assay(se),
  group=se$dex,
  samples=as.data.frame(colData(se))[, c('cell', 'dex')])
y.all <- calcNormFactors(y.all)

## Annotate genes
y.all$genes <- data.frame(
  ensg_id=rownames(y.all),
  entrez_id=mapIds(org.Hs.eg.db, rownames(y.all), 'ENTREZID', 'ENSEMBL'),
  symbol=mapIds(org.Hs.eg.db, rownames(y.all), 'SYMBOL', 'ENSEMBL'),
  length=sum(width(rowRanges(se))),
  stringsAsFactors=FALSE)
```

Here is an overview of the data we are working with

`r knitr::kable(y.all$samples)`

## Data Analysis

We want to identify the genes (and genesets) regulated under `dex` treatment.

We'll use the following design:

```{r design-setup, results='hide'}
des <- model.matrix(~ 0 + group + cell, y.all$samples)
colnames(des) <- sub('group', '', colnames(des))
des
```

`r knitr::kable(as.data.frame(des))`

To test for the general affect of dexamethasone treatment independant of
cell-type, we would test the following contrast:

```{r contrast-setup}
cm <- makeContrasts(dex=trt - untrt, levels=des)
cm
```

<div class="note">
Many of the GSEA methods provided by limma support arbitrarily
complex experimental designs. For instance, we can consider the `cell` factor
to be a random effect (as opposed to the fixed effect we've setup above) and use
limma's `duplicateCorrelation` functionality to model this, like so:

```{r random-effects, eval=FALSE}
## This code isn't runnable now. It is included here to provide an
## example of the more exotic models that multiGSEA can support by
## virtue of leveraging limma internally.
des.re <- model.matrix(~ 0 + group, y.all$samples)
vm.re <- voom(y, des.re)
corfit <- duplicateCorrelation(vm.re, vm.re$design, block=y$samples$cell)
vm.re <- voom(y, des.re, plot=TRUE, block=y$samples$cell,
              correlation=corfit$consensus)
```
</div>

### Differential Gene Expression

In this section, we first show you the straightforward analysis you would do
if you were only testing for differential gene expression.

We first filter out genes from the dataset that are either poorly annotated
(ie. no entrez ID) or are lowly expressed. We consider a gene to be lowly
expressed if it's not expressed above 1 cpm in at least 4 experiments (because
this dataset has n=4 replication between dex treatments).

Note that the rowname/gene identifiers are currently Ensemble Gene Identifiers 
(eg. `ENSG00000119203`), and we convert these to EntrezIDs for easier downstream
analysis.

```{r gene-filter, fig.width=5, fig.height=4.5}
keep <- !is.na(y.all$genes$entrez_id) &
  !duplicated(y.all$genes$entrez_id) &
  rowSums(cpm(y.all) >= 1) >= 4
y <- y.all[keep,,keep.lib.sizes=FALSE] %>%
  calcNormFactors() %>%
  set_rownames(., .$genes$entrez_id) ## using entrez identifiers
vm <- voom(y, des, plot=FALSE)
```

Now that we have our expression (`vm`), design (`vm$design`) and contrast (`cm`)
data in hand, we can perform our differential expression analyss.

```{r dge-analysis}
fit <- lmFit(vm, vm$design) %>%
  contrasts.fit(cm) %>%
  eBayes
tt.dex <- topTable(fit, 'dex', n=Inf, sort.by='none')
```

### Gene Set Enrichment Analysis

Given that we now have all of the pieces of data required for a differential
expression analysis, performing GSEA is trivial using the `multiGSEA` wrapper
function. We simply need to now define (1) the battery of gene sets we want to
test against, and (2) the GSEA methods we want to explore.

#### Gene Sets to Test

The *multiGSEA* package provides a `GeneSetDb` class, which is a new container
(insted of `GSEABase::GeneSetCollection`) to house collections of genesets. You
can easily create a `GeneSetDb` object from a custom set of gene sets you have
on hand, but the package also provides convenience wrappers to generate 
`GeneSetDb` objects from popular gene annotation resources such as
[MSigDB][msigdb], [PANTHER][pantherdb], etc.

We'll use the `getMSigGeneSetDb` convenience function provided by the
*multiGSEA* package to simply get the hallmark geneset collection.

[msigdb]: http://software.broadinstitute.org/gsea/msigdb/
[hallmark]: http://software.broadinstitute.org/gsea/msigdb/collections.jsp#H
[c2]: http://software.broadinstitute.org/gsea/msigdb/collections.jsp#C2
[c5]: http://software.broadinstitute.org/gsea/msigdb/collections.jsp#C5
[c7]: http://software.broadinstitute.org/gsea/msigdb/collections.jsp#C7
[pantherdb]: http://pantherdb.org
[pgoslim]: http://www.pantherdb.org/panther/ontologies.jsp

```{r build-gdb, eval=!exists('gdb')}
library(multiGSEA)
gdb <- getMSigGeneSetDb('h',  'human')
```

You can view a table of the gene sets defined inside a `GeneSetDb` (`gdb`)object
via its `geneSets(gdb)` accessor:

```{r geneSets-accessor}
geneSets(gdb) %>%
  head %>%
  select(1:4) %>%
  knitr::kable()
```

For more details on creating and manipluating `GeneSetDb` objects, please jump
the to *[The GeneSetDb Class](#the-genesetdb-class)* section.

#### Running multiGSEA

Performing multiple gene set enrichment analyses over your contrast of interest
simply requires you to provide a `GeneSetDb` object along with your data and an
enumeration of the methods you want to use in your analysis.

The call to `multiGSEA` will perform these analyses and return a
`MultiGSEAResult` object which you can then use for downstream analysis.

```{r run-multi-GSEA, eval=!exists('mg.dex'), cache=TRUE}
mg.dex <- multiGSEA(gdb, vm, vm$design, cm[, 'dex'],
                    methods=c('camera', 'fry', 'goseq'),
                    ## GSEA specific parameters
                    feature.max.padj=0.01, feature.min.logFC=1,
                    feature.bias=setNames(vm$genes$length, rownames(vm)),
                    inter.gene.cor=0.01)
```

#### Implicit Differential Expression

First, let's note that in addition to running a plethora of GSEA's over our data
we've also run a standard differential expression analysis. If you've passed
a `matrix`, `ExpressionSet` or `EList` into `multiGSEA`, a *limma*-based
`lmFit  %>% (eBayes|treat) %>% (topTable|topTreat)` pipeline was run. If a
`DGEList` was passed, then `multiGSEA` utilizes the *edgeR*-based
`glmQLFit %>% (glmQLFTest | glmTreat) %>% topTags` pipeline.

Either way, the results of the internally run differential expression analysis
is accessible via a call to `logFC` function on the `MultiGSEAResult` object:

```{r logFC-results, results='asis'}
lfc <- logFC(mg.dex)
lfc %>%
  select(symbol, entrez_id, logFC, CI.L, CI.R, t, pval, padj) %>%
  head() %>%
  knitr::kable(digits=3)
```

We can confirm that the statistics generated internally in multiGSEA mimic our
explicit analysis above by verifying that the t-statistics generated by both
approaches are identical.

```{r compare-dge-t-stats}
comp <- select(tt.dex, entrez_id, symbol, logFC, t) %>%
  inner_join(lfc, by=c('entrez_id'), suffix=c('.tt', '.mg'))
all.equal(comp$t.tt, comp$t.mg)
```

```{r plot-dge-t-stats, fig.width=4.5, fig.height=4, eval=FALSE, include=FALSE}
ggplot(comp, aes(t.tt, t.mg)) +
  geom_point(alpha=0.7) +
  geom_abline(slope=1, intercept=0, color='red') +
  xlab("t-statistics (topTable)") +
  ylab("t-statistics (multiGSEA)") +
  ggtitle("t-statistic comparison")  
```

The internally performed differential expression analysis within the `mulitGSEA`
call can be customized almost as extensively as an explicitly performed analysis
that you would run using limma or edgeR by sending more paramaters through
`multiGSEA`'s `...` argument.

See the
*[Custom Differential Expression](#custom-differential-expression)*
section further in the vignette as well as the help available in
`?calculateIndividualLogFC` (which is called inside the `multiGSEA` function)
for more information.

#### Explicit GSEA

We also have the results of all the GSEA analyses specified in our `methods`
parameter.

```{r mg-res}
mg.dex
```

### Exploring Results

GSEA results can be examined interactively via the command line, or via a shiny
app. You can use the `resultNames` function to find out what results exist
within the `MultiGSEAResult` object:

```{r resultnames}
resultNames(mg.dex)
```

The individual gene set statistics generated by each method are available via
the `result` function (or several can be returned with `results`:

```{r goseq-res}
go.res <- result(mg.dex, 'goseq')
go.up <- result(mg.dex, 'goseq.up')
go <- results(mg.dex, c('goseq', 'goseq.up'))
```

You can also list the members of a geneset and their individual differential
expression statistics for the contrast under test using the `geneSet` function.

```{r geneset-result}
tnfa <- geneSet(mg.dex, 'h', 'HALLMARK_TNFA_SIGNALING_VIA_NFKB')
tnfa %>%
  select(symbol, entrez_id, logFC, pval, padj) %>%
  head %>%
  knitr::kable(digits=3)
```

<div class="note">
The results provided in the table generated from a call to `geneSet` are
independant of GSEA method.
</div>

#### Plotting

*multiGSEA* provides a number of interactive plotting facilities to explore the
enrichment of a single geneset under the given contrast.

**Boxplot**

```{r iplot-tnfa-box, fig.width=4, fig.height=3}
iplot(mg.dex, 'h', 'HALLMARK_TNFA_SIGNALING_VIA_NFKB',
      type='boxplot', value='t')
```

**Density**

```{r iplot-tnfa-density, fig.width=4, fig.height=3}
iplot(mg.dex, 'h', 'HALLMARK_TNFA_SIGNALING_VIA_NFKB',
      type='density', value='t')
```

**Barcode Plot**

You might also like limma's barcode plot (which hasn't been made interactive
yet):

```{r plot-barcode, fig.width=5, fig.height=3.5}
plot(mg.dex, 'h', 'HALLMARK_TNFA_SIGNALING_VIA_NFKB',
     type='barcode', value='t')
```

#### Interactive Exploration

For now, the interactivity is provided by the sister *multiGSEA.shiny* package,
which is accessible via http://go.gene.com/multiGSEA and in
[subversion here][mgssvn].

<div class="important">
Please only advertise the http://go.gene.com/multiGSEA URL for
this webapp. The exact application URL this redirects to may change over time.
</div>

You need to first save the `MultiGSEAResult` object to your filesystem, like so:

```{r save-rds, eval=FALSE}
saveRDS(mg.dex, 'mgdex.rds')
```

Then upload the file to [the shiny application][mgsweb].

You can [download a serialized version of `mg.dex` here][mgdex] to play with the
app now, but you'll probably have more fun poking around the application using
results from a dataset more important to you.

<div class="note">
For now this is the mechanism you need to use to allow your collaborators to
explore the results. It is unfortunately too manual.

Eventually we will:

  * Have this more easily automated within ExpressionPlot; and
  * Allow the analyst to simply explore these results within an active R session
    without having to save the object and manually launch the shiny application.
    A simple call to `explore(mg.dex)` will launch the app with the results
    preloaded for exploration.
</div>

[mgsweb]: http://go.gene.com/multiGSEA
[mgssvn]: http://resscm.gene.com/bioinfo/projects/R/trunk/multiGSEA.shiny/
[mgdex]: http://resdev.gene.com/people/lianogls/Rpkgs/multiGSEA/Rpkg/vignettes/johnnycache/mg-dex.rds

# The GeneSetDb Class

The GeneSetDb class is a new container to store genesets.

  * Why not `GSEABase::GeneSetCollection`?
  * Constructing custom `GeneSetDb`s:
      * From a `list` of `lists` 
      * From a `data.frame`
  * `gdb <- conform(gdb, vm)`
      * Identifies the features you've measured in the experiment `vm`
      * Appropriately sets the `universe` of tests that require it
      * It's "non-destructive", you can reconform the same `GeneSetDb` to
        another dataset without loss of information (geneset membership)
      * Use of `featureIdMap`

## Building a GeneSetDb

The multiGSEA package provides conveneince funcitons to fetch genesets from many
sources and convert them into a GeneSetDb object. These sources include:

* MSigDB
* PANTHER (pathways and GOSLIM)
* More ...

## Active vs Inactive Gene Sets

After a `GeneSeDb` is conformed to a target expression object `E`,
individual gene sets become "active" or "inactive" depending on the number
of features per gene set are found in `E`. For instance, if `E`
does not have measurements for at least `conform(, min.gs.size)`
features for a given gene set, that gene set will be flagged as inactive.

## Mapping of gene set featureIds to target expression containers

It may be that the IDs used in a gene set collection are different from the
ones used as the rownames of your expression cotainer. For instance, the IDs
used for a given gene set collection in the \code{GeneSetDb} might be
Ensembl gene identifiers, but the rownames of the expression object migth
be Entrez ID. This is where the \code{mapping} parameter becomes useful.

TODO: document the GeneSetDb ID mapping more thoroughly.

# Costumizing Analyses

The `...` in `multiGSEA(gdb, x, design, coef, methods, ...)` are you friend.

## Custom Differential Expression

Works on `EList` and `DGEList` (using edgeR's quasilikihood framework)
    * `robust.fit`
    * `robust.eBayes`

## Custom GSEA


# Extending multiGSEA

## Adding new GSEA methods

There are three funtions you need to provide in order to add a new method `XXX`
to the multi `multiGSEA` package:

  * `validate.inputs.XXX`
  * `do.XXX`
  * `summarize.XXX`

### Validate Inputs

In general, it seems that GSEA methods require "just" a vector of statistics
that enrichment can be run over, or a full representation of an experiment and
a specific contrast to test.

If your method simply requires a vector of statistics, like
\code{limma::geneSetTest}, then alias your `validate.inputs.XXX` function to
`multiGSEA::.validate.inputs.logFC.only`. If the method is of the latter
variety, like \code{limma::camera} then alias your `validate.inputs.XXX` method
to `multiGSEA:::..validate.inputs.full.design`, ie:

    validate.inputs.my_new_method <- multiGSEA:::..validate.inputs.full.design

### Do the analysis

The `do.XXX` method just support the following signature:

    do.my_new_method(gsd, x, design, contrast, ...)

If your method only requires a vector of statistics, `design` and `contrast`
your method can simply ignored `design` and `contrast` (or, you can let your
`...` consume them).

### Summary of results

The `sumamrize.XXX` method should return a data.table that (minimally) had
a `collection` and `id` column followd by relevant information/statistics
produced by your method for each geneset that was tested.

# Scratch

## Background reading for Gene Set Enrichment Analyses


Overview of The Broad's GSEA:
http://baderlab.org/CancerStemCellProject/VeroniqueVoisin/AdditionalResources/GSEA


## Himes airway paper

Quotes from the Himes et al. paper:

* Differential Gene Expression:

    > Overall, 316 genes were significantly differentially expressed after correcting for false discovery rate by the Benjamini-Hochberg [23] approach [Figure 1A, Table S3]. Table 1 contains the genes with Q-value <1E-10

* GSEA:

    > Gene set enrichment analysis using the NIH DAVID tool [28] identified various Gene Ontology and other annotation categories that were overrepresented by the 316 genes. The top six gene functional annotation clusters (enrichment scores >3) had terms related to: glycoprotein/extracellular matrix, vasculature development, circulatory system process, response to nutrients, thrombospondin type-1, and response to hormone stimulus terms [Table S4]. Other clusters among the 19 with enrichment scores >1.5 that may be relevant to lung disease included lung development, regulation of cell migration, and extracellular matrix organization.
